\chapter{Discussion}
\label{c:discussion}

In this chapter, we first describe our contributions and then discuss possible directions for future work.

\section{Contributions}

We list below the contributions achieved with this dissertation.

First and more important, we developed a mechanisation using the \textit{Rocq Prover} and the \textit{Autosubst} library of the following systems introduced in this work:
\begin{enumerate}
\item the multiary $\Lam$-calculus (system $\LamM$);
\item the canonical subsystem of $\LamM$;
\item the canonical $\Lam$-calculus (system $\LamV$).
\end{enumerate}
Taking the formalisations of these systems on the proof assistant, we also obtained computer verified proofs for results such as:
\begin{enumerate}
\item subject reduction for systems $\LamM$ and $\LamV$;
\item isomorphism between the canonical subsystem of $\LamM$ and system $\LamV$;
\item conservativeness of $\LamV$ over $\LamM$;
\item isomorphism between the simply typed $\Lam$-calculus and system $\LamV$.
\end{enumerate}

Second, we gave an exhaustive definition for the concept of subsystem, separating two isomorphic representations of the canonical subsystem of $\LamM$.
This helped us clarify the loose idea of subsystem and simplify some of the result proven (for example, using the self-contained system $\LamV$ for proving the theorem of conservativeness).
From this idea, we could even propose a standard aproach to formalise any subsystem.

Third, through this document, a detailed exposition of the mechanised systems and proofs in the \textit{Rocq Prover} along with some digressions over formalisation choices.

Fourth and last, we layed out a modularised formalisation that helped us enjoy the automated tactics of the \textit{Rocq Prover}.
This way, many of the provided proofs were obtained automatically.

\section{Further work}

Why use a outdated library for mechanising binders? What about \textit{Autosubst 2}?

\begin{itemize}
\item Could there be more automation? ...?
\item Because of the modularity and of the \textit{Autosubst} library we have the facility to enrich our typing systems (ex: SystemF?).
\item Many more mechanisations of metatheory?
\end{itemize}

\begin{comment}
Mechanisations in relation with the formalisations on the paper.
\begin{itemize}
\item Some ideas for the metatheory formalised come from attempts of mechanisations.
\item The metatheory mechanised does not correspond exactly to the formalised in the literature.
  \subitem An example: the polymorphic definition for system  $\LamM$.
\end{itemize}

Further work?
\begin{itemize}
\item Because of the modularity and of the \textit{Autosubst} library we have the facility to enrich our typing systems (ex: SystemF?).
\item Could there be more automation?  
\item Why use a outdated library for mechanising binders? What about \textit{Autosubst 2}?
\item Obstacles on the mechanisation of a non standard substitution operation.
\item SSreflect style proofs for \textit{Rocq Prover}.
\end{itemize}
\end{comment}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "../dissertation"
%%% End:
