\chapter*{Acknowledgments}

I feel blessed to write this, as I contemplate so many people that I have to thank, as I see so many people that I love and love me back.

First and foremost, I must thank my family for allowing me to pursue what I love and for unconditionally supporting my passions.

Second, I want to deeply thank my girlfriend, Inês, for being my companion throughout this year of confusion, tiredness and grumpiness.
I still owe you a complete explanation of what $\lambda$-calculus is.

Third, I want to thank all my friends...should I name them?
To Nando, Marco and David, as you are a second kind of girlfriend to me.
To Gaspar, Rodrigo and Kiko, as you have known me for a long time and part of who I am today is because of you.
To Pepa, as you may not understand how much of a help you were to me throughout this year.
To Gui, Zé, Bruno, Sofia, Manada, Gonçalo, Jason, Jonny and to every algebra class that made us closer.
To Faustino, Ricardo, Guilherme and Filipa, for making me feel more at home in Braga.
To my long time friends, Couves e Beterrabas, for being with me in such an important and playful time in my life.
To my CVX group, VitaminaC, for walking with me.
To CREU, Missão País, Fonte da Prata and to every friend that I have made in those special places - a piece of my heart lies with each of you.

Fourth and last, I want to thank everyone who I feel had a part in my logical education, from Rola and Rita to Professor Filipe and Professora Alice Viveiros.
And, of course, to my kind supervisors.
Thank you, Professor Luís, for your patience and for your focused, optimistic and close help throughout this joyful process.
Thank you, Professor José Carlos, for your enthusiasm and for every exciting conversation we have shared.

A last special thanks to  the Research Centre of Mathematics of the University of Minho (CMAT) for funding this dissertation through the CMAT Research Grant UIDB/00013/2020.

% To everyone that casually asked what I was studying in my dissertation and kept their attention throughout my whole explanation - because attention without distraction is a prayer - I thank you.

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "../dissertation"
%%% End:
